---
output:
  html_document:
    css: styles.css
---

<br>

## Forecasting Weekly Retail Demand at Walmart
**Darian Miranda Olivas**

## Overview
End to end machine learning workflow for forecasting weekly Walmart retail sales at the store-department level using historical sales data.

## Model Performance
Final model achieved **8.4% WAPE** on the test dataset, demonstrating strong predictive performance on unseen future demand.

## Data Source
This analysis uses the Walmart Recruiting – Store Sales Forecasting dataset made available through a Kaggle competition hosted by Walmart Inc. Access the data [here](https://www.kaggle.com/competitions/walmart-recruiting-store-sales-forecasting/data).

<br>

#### Data Integration
**Purpose:** Prepare raw data into a consistent format ready for analysis.
<br>
**Features Created:**
<br>
`week_dates`
<br>
`store_dept`

```{r data_integration, echo=TRUE, message=FALSE, warning=FALSE}

#====== Set seed for reproducibility ======

RNGkind("L'Ecuyer-CMRG")
set.seed(2067205) ##seed number was randomly typed into keyboard





#====== Load in necessary packages ======

library(pacman) ##for bulk package loading

p_load(finetune, ##for efficient tuning 
        janitor, ##for consistent feature names
        modeltime, ##for autoregressive forecasting
        skimr, ##for inspecting data
        slider, ##for rolling window statistics
        tidymodels, ##for modeling
        tidyverse, ##for everything
        timeDate, ##for holidays
        vip, ##for visualizing feature importance
        xgboost) ##for XGBoost model





#====== Read in data as tibbles (tbls) ======

sales_tbl <- read_csv("data.csv/sales.csv") ##raw sales data

features_tbl <- read_csv("data.csv/features.csv") ##external economic and price mark down data

stores_tbl <- read_csv("data.csv/stores.csv") ##store metadata





#====== Join tbls and update ID features ======

base_tbl <- sales_tbl %>%
  left_join(features_tbl,
            by = c("Date", "IsHoliday", "Store")) %>% 
  left_join(stores_tbl,
            by = "Store") %>%
  clean_names(case = "snake") %>% ##snake case transforms all names to lowercase characters and underscores for any non characters
  mutate(
    
    week_dates = as.Date(date),
    store_dept = interaction(store, dept, drop = TRUE)
    
         ) %>%
  select(-date)

```
<br>

#### Raw Data Validation
**Purpose:** To ensure the raw data is suitable feature engineering.

```{r raw_data_validation, echo=TRUE, message=FALSE, warning=FALSE}

#====== Inspect data ======

glimpse(base_tbl)





#====== Check for missing values ====

colSums(is.na(base_tbl))

##Result is that there are missing values detected in mark_down features





#====== Check number of stores and departments =====

n_distinct(base_tbl$store)


n_distinct(base_tbl$dept)

##There are 45 stores and 81 departments





#====== Check statistics for key numeric variables =====

base_tbl %>%
  select(store, dept, weekly_sales) %>%
  summary()

##All store and dept ids are positive
##Negative weekly sales values exist in the data, assumed loss and not data error





#====== Check for duplicates ======

anyDuplicated(base_tbl)

##Result is that there are no duplicated rows

```
<br>

#### Calendar Features
**Purpose:** To capture calendar driven retail demand.
<br>
**Features Created**
<br>
`week`, `month`, `quarter`, `year`
<br>
`week_of_{month, quarter}`
<br>
`is_quarter_{start, end}`
<br>
`week_{sin, cos}`

```{r calendar_fts, echo=TRUE, message=FALSE, warning=FALSE}

#====== Create calendar features ======

calendar_fts_tbl <- base_tbl %>%
  distinct(week_dates) %>%
  mutate(
    
    #---- basic calendar features ----
    week    = week(week_dates),
    month   = month(week_dates),
    quarter = quarter(week_dates),
    year    = year(week_dates),
    
    #---- week index features ----
    week_of_month    = ceiling(day(week_dates) / 7), 
    week_of_quarter  = ceiling(yday(week_dates) / 7) - ceiling(yday(floor_date(week_dates, "quarter")) / 7), 
    
    #---- quarter edge features ----
    is_quarter_start = as.integer(week_of_quarter <= 2), ##quarter start defined as first 2 weeks
    is_quarter_end   = as.integer(week_of_quarter >= 12), ##quarter end defined as last 2 weeks
    
    #---- continuous seasonality features ----
    week_sin = sin(2 * pi * week / 52), ##sin and cos functions replicate cyclical nature of time
    week_cos = cos(2 * pi * week / 52)
    
  )

```
<br>

#### Holiday/Event Features
**Purpose:** To capture holiday/event driven retail demand.
<br>
**Features Created**
<br>
`is_{holiday/event}`
<br>
`back_to_school_ramp`
<br>
<br>
holiday/event ∈
<br>
{`valentines`, `independence_day`, `halloween`, `christmas`, `new_years_eve`, `super_bowl`, `mothers_day`, `memorial_day`, `fathers_day`, `labor_day`, `thanksgiving`, `black_friday`, `easter`}

```{r holiday_event_fts, echo=TRUE, message=FALSE, warning=FALSE}

#====== Create tbl with all days in training data ======

days_tbl <- tibble(day_dates = seq(min(base_tbl$week_dates), 
                                   max(base_tbl$week_dates), 
                                   by = "day"))





#====== Create a vector of years in training data ======

years <- seq(year(min(days_tbl$day_dates)), 
             year(max(days_tbl$day_dates))) 





#====== Get dates for Easter in training data ======

easter_dates <- as.Date(Easter(years)) ##use package since set by unique Paschal full moon





#====== Create a tbl with a feature for each holiday/event ======

holiday_event_tbl <- days_tbl %>% 
  mutate(
    
    #---- fixed date holiday features ----
    is_valentines     = month(day_dates) == 2  & mday(day_dates) == 14, 
    is_independence_day  = month(day_dates) == 7  & mday(day_dates) == 4,
    is_halloween      = month(day_dates) == 10 & mday(day_dates) == 31,
    is_christmas      = month(day_dates) == 12 & mday(day_dates) == 25,
    is_new_years_eve  = month(day_dates) == 12 & mday(day_dates) == 31,
    
    #---- pattern based holiday features ----
    is_super_bowl = ##1st sun of feb
      wday(day_dates) == 1 & 
      month(day_dates) == 2 & 
      mday(day_dates) <= 7,
   
    is_mothers_day    = #2nd sun of may
      wday(day_dates) == 1 &
      month(day_dates) == 5 &
      mday(day_dates) > 7 &
      mday(day_dates) <= 14,
      
     is_memorial_day   = ##last mon of may
      wday(day_dates) == 2 & 
      month(day_dates) == 5 & 
      mday(day_dates) >= 25,

    is_fathers_day    = #3rd sun of jun
      wday(day_dates) == 1 & 
      month(day_dates) == 6 & 
      mday(day_dates) > 14 & 
      mday(day_dates) <= 21,

    is_labor_day      = ##1st mon of sep
      wday(day_dates) == 2 & 
      month(day_dates) == 9 & 
      mday(day_dates) <= 7,
      
    is_thanksgiving = ##4th thu of nov 
      wday(day_dates) == 5 & 
      month(day_dates) == 11 & 
      mday(day_dates) >= 22 & 
      mday(day_dates) <= 28,

    is_black_friday = ##fri after thanksgiving
      wday(day_dates) == 6 &
      month(day_dates) == 11 &
      mday(day_dates) >= 23 &
      mday(day_dates) <= 29,
    
    is_easter = day_dates %in% easter_dates,
   
   #---- back to school ramp feature ----
    back_to_school_ramp = case_when( ##back to school happens at different times across the United States
      day_dates %in% 213:220 ~ 1, ##early back to school
      day_dates %in% 227:241 ~ 2, ##peak back to school
      day_dates %in% 248:255 ~ 1, ##late back to school
      TRUE ~ 0)

   )





##wday() function from Lubridate package acts inconsistently inside of mutate() function
##for a stable solution I define a weekday indexing function

#====== Define a weekday indexing function ======

wday_function <- function(day_dates) { 
  (wday(day_dates)) %% 7 + 1 ##first entry of data is 2010-02-05, so index Friday as 7
}





#====== Map holiday daily data to respective sales weeks ======

holiday_event_fts_tbl <- holiday_event_tbl %>%
  ##use algorithmic approach to round up to Friday because time round functions default to Sunday
  mutate(week_dates = day_dates +  7 - wday_function(day_dates)) %>% 
  group_by(week_dates) %>% 
  summarise(
    
    across(starts_with("is_"),
           ~ any(.x, na.rm = TRUE)), ##inserts TRUE for a holiday feature if any day of that sales week is a holiday
    
    back_to_school_ramp = max(back_to_school_ramp, na.rm = TRUE),
    
    .groups = "drop"
    
    )

```
<br>

#### Holiday Timing Features
**Purpose:** 
<br>
**Features Created**
<br>
`lead_is_{holiday/event}_1`
<br>
`lead_is_{major_holiday}_{2,3}`
<br>
`lead_is_christmas_4`
<br>
`lag_is_{holiday/event}_1`
<br>
`lag_is_christmas_{2,3,4}`
<br>
<br>
holiday/event ∈ 
<br>
{`valentines`, `independence_day`, `halloween`, `christmas`, `new_years_eve`, `super_bowl`, `mothers_day`, `memorial_day`, `fathers_day`, `labor_day`, `thanksgiving`, `black_friday`, `easter`}
<br>
<br>
major_holiday ∈ 
<br>
{`valentines`, `independence_day`, `halloween`, `christmas`, `new_years_eve`, `super_bowl`, `thanksgiving`, `easter`}

```{r holiday_timing_fts, echo=TRUE, message=FALSE, warning=FALSE}

#====== Create a vector of major holidays/events ======

##major holidays are defined as events where multi day preparation or extended demand shifts are expected, single day demand spikes (like Black Friday) are excluded

major_holidays <- c("valentines",
                    "independence_day", 
                    "halloween", 
                    "christmas", 
                    "new_years_eve", 
                    "super_bowl", 
                    "thanksgiving", 
                    "easter")





##lead features here are safe because future holiday indicators are available at prediction time and do not introduce data leakage

#====== Create holiday specific lead features ======

lead_holiday_fts_tbl <- holiday_event_fts_tbl %>%
  select(week_dates, starts_with("is")) %>%
  arrange(week_dates) %>% 
  mutate(
    
     #---- 1 week lead for all holidays ----
     across(starts_with("is"),
           ~ lead(.x, 1), ##recall data is weekly so 1 period lead
           .names = "lead_{.col}_1"),
     
     #---- 2-3 weeks for major holidays only ----
     across(starts_with("is") & contains(major_holidays),
            list(
              `2` = ~ lead(.x, 2),
              `3` = ~ lead(.x, 3)
            ),
            .names = "lead_{.col}_{.fn}"),
     
     #---- 4 week lead for Christmas only ----
     across(is_christmas,
            ~ lead(.x, 4),
            .names = "lead_{.col}_4")
    
    ) %>%
    select(week_dates, contains("lead_is_"))





#====== Create holiday specific lag features ======

lag_holiday_fts_tbl <- holiday_event_fts_tbl %>%
  select(week_dates, starts_with("is")) %>%
  arrange(week_dates) %>% 
  mutate(
    
    #---- 1 week lag for all holidays ----
    across(starts_with("is"),
           ~ lag(.x, 1),
           .names = "lag_{.col}_1"),
  
    #---- 2-4 week lags for Christmas only ----
    across(is_christmas,
      list(
        `2` = ~ lag(.x, 2),
        `3` = ~ lag(.x, 3),
        `4` = ~ lag(.x, 4)
        ),
      .names = "lag_{.col}_{fn}")
    
    ) %>%
  select(week_dates, contains("lag_is_"))

```
<br>

#### Demand Lag Features
**Purpose:** To capture dependence of demand on recent demand history.
<br>
**Features Created**
<br>
`lag_weekly_sales_{1,2,4,13,26}`

```{r lag_demand_fts, echo=TRUE, message=FALSE, warning=FALSE}

#====== Create sales lag features ======

lag_demand_fts_tbl <- base_tbl %>%
  select(store, dept, week_dates, weekly_sales) %>%
  arrange(store, dept, week_dates) %>%
  group_by(store, dept) %>%
  mutate(
    
    across(weekly_sales,
           list(
             `1`  = ~ lag(.x, 1),
             `2`  = ~ lag(.x, 2),
             `4`  = ~ lag(.x, 4),
             `13` = ~ lag(.x, 13),
             `26` = ~ lag(.x, 26)
           ),
           .names = "lag_{.col}_{.fn}")
    
    ) %>%
    ungroup() %>%
    select(store, dept, week_dates, contains("lag_weekly_sales_"))

```
<br>

#### Demand Level And Variability
**Purpose:** To capture level and variability of retail demand in recent history.
<br>
**Features Created**
<br>
`roll_mean_{4,8,13,26}`
<br>
`roll_sd_{4,8,13,26}`
<br>
`is_demand_shock`

```{r level_variability_demand, echo=TRUE, message=FALSE, warning=FALSE}

#====== Create demand level features ======

level_demand_fts_tbl <- lag_demand_fts_tbl %>%
  select(store, dept, week_dates, lag_weekly_sales_1) %>%
  arrange(store, dept, week_dates) %>%
  group_by(store, dept) %>%
  mutate(
    
    roll_mean_4  = slide_dbl(lag_weekly_sales_1, mean, .before = 3, .complete = TRUE), #.before = 3 means use current period and 3 before, gives 4 week window
    roll_mean_8  = slide_dbl(lag_weekly_sales_1, mean, .before = 7, .complete = TRUE),
    roll_mean_13 = slide_dbl(lag_weekly_sales_1, mean, .before = 12, .complete = TRUE),
    roll_mean_26 = slide_dbl(lag_weekly_sales_1, mean, .before = 25, .complete = TRUE),
    
    ) %>%
  ungroup() %>%
  select(store, dept, week_dates, contains("roll_mean_"))





#====== Create demand variability features ======

variability_demand_fts_tbl <- lag_demand_fts_tbl %>%
  select(store, dept, week_dates, lag_weekly_sales_1) %>%
  left_join(level_demand_fts_tbl %>%
      select(store, dept, week_dates, roll_mean_13),
      by = c("store", "dept", "week_dates")
      )%>%
  arrange(store, dept, week_dates) %>%
  group_by(store, dept) %>%
  mutate(
    
    roll_sd_4  = slide_dbl(lag_weekly_sales_1, sd, .before = 3, .complete = TRUE),
    roll_sd_8  = slide_dbl(lag_weekly_sales_1, sd, .before = 7,  .complete = TRUE),
    roll_sd_13 = slide_dbl(lag_weekly_sales_1, sd, .before = 12, .complete = TRUE),
    roll_sd_26 = slide_dbl(lag_weekly_sales_1, sd, .before = 25, .complete = TRUE),
    
    #---- demand shock feature ----
    is_demand_shock = abs(lag_weekly_sales_1 - roll_mean_13) > 2 * roll_sd_13 ##2 sd comes from statistical anomaly rule
    
    ) %>%
  ungroup() %>%
  select(store, dept, week_dates, contains(c("roll_sd_", "is")))

```
<br>

#### Demand Momentum Features
**Purpose:** To capture directional movement and trend persistence in demand.
<br>
**Features Created**
<br>
`avg_same_week_last_year`
<br>
`growth_yoy`
<br>
`ratio_demand_short_medium`
<br>
`roll_slope_8` 
<br>
`ewma_6`

```{r momentum_demand_fts, echo=TRUE, message=FALSE, warning=FALSE}

#====== Load or create demand momentum feature layer ======

version <- "v1"
momentum_demand_fts_path <- paste0("r/momentum_demand_fts_tbl_",  version, ".rds")


if (file.exists(momentum_demand_fts_path)) {
  
  message("Loading existing momentum feature layer...")
  momentum_demand_fts_tbl <- readRDS(momentum_demand_fts_path)
  
} else {

  
  
  
  
#====== Define a function that calculates slope ======

slope_function <- function(x) {
  
  time_index <- seq_along(x)
  lm_fit <- lm(x ~ time_index) ##regress sales on time
  
  coef(lm_fit)[2]
  
        }





#====== Define a function that calculates exponential weighted moving average (EWMA) ======

ewma_function <- function(x, alpha) {
  ewma_value <- x[1]
  
  for (i in 2:length(x)) {
    ewma_value <- alpha * x[i] + (1 - alpha) * ewma_value ##EWMA basic formula
  }
  
  ewma_value
}



    

#====== Create demand momentum features ======

momentum_demand_fts_tbl <- base_tbl %>%
  select(store, dept, week_dates, weekly_sales) %>%
   left_join(lag_demand_fts_tbl %>%
               select(store, dept, week_dates, lag_weekly_sales_1),
             by = c("store", "dept", "week_dates")
             ) %>%
  left_join(level_demand_fts_tbl %>%
              select(store, dept, week_dates, roll_mean_4, roll_mean_13),
            by = c("store", "dept", "week_dates")
            ) %>%
  arrange(store, dept, week_dates) %>% 
  group_by(store, dept) %>%
  mutate(
    
    #---- year over year growth feature ----
    growth_yoy = (lag_weekly_sales_1 - lag(weekly_sales, 53)) / pmax(lag(weekly_sales, 53), 1), ##prevent division by 0 with pmax(x,1)
    
    
    
    #---- short term vs medium term demand ratio feature ----
    ratio_demand_short_medium = roll_mean_4 / pmax(roll_mean_13, 1),
    
    
    
    #---- same week last year average feature ----
    avg_same_week_last_year = slide_dbl(
      lag(weekly_sales, 52), ##1 year prior
      mean,
      .before = 3, ##use a 4 week window since singular week sales can be noisy
      .complete = TRUE
      ),
    
    
    
    #---- rolling slope feature ----
    roll_slope_8 = slide_dbl(
      .x      = lag_weekly_sales_1,
      .before = 7,
      .f      = ~ slope_function(.x),
      .complete = TRUE
      ),
    
    
    
    #---- EWMA feature ----
    ewma_6 = slide_dbl( 
      .x      = lag_weekly_sales_1,             
      .before = 5,  
      .f      = ~ ewma_function(.x, alpha = 0.3),
      .complete = TRUE
      ),
    
    ) %>%
  ungroup() %>%
  select(store, dept, week_dates, contains(c("growth", "ratio", "avg", "roll_slope_", "ewma_")))





#====== Save demand momentum features tbl ======

saveRDS(momentum_demand_fts_tbl, momentum_demand_fts_path)

}

```
<br>

#### Economic Features
**Purpose:** To capture dependence of demand on external economic forces.
<br>
**Features Created**
<br>
`lag_cpi_{4,12}`
<br>
`lag_unemployment_{4,12}`
<br>
`trend_{cpi,unemployment}`

```{r economic_fts, echo=TRUE, message=FALSE, warning=FALSE}

#====== Create economic features ======

economic_fts_tbl <- base_tbl %>%
  distinct(week_dates, .keep_all = TRUE) %>%
  select(week_dates, cpi, unemployment) %>%
  arrange(week_dates) %>%
  mutate(

    #---- 4 week economic lag features ----
    lag_cpi_4  = lag(cpi, 4),
    lag_unemployment_4 = lag(unemployment, 4),

    #---- 12 week economic lag features ----
    lag_cpi_12  = lag(cpi, 12),
    lag_unemployment_12 = lag(unemployment, 12),

    #---- economic momentum features ----
    trend_cpi = lag_cpi_4 - lag_cpi_12,
    trend_unemployment = lag_unemployment_4 - lag_unemployment_12

  ) %>%
  select(week_dates, contains(c("lag", "trend")))

```
<br>

#### Mark Down Features
**Purpose:** To capture dependence of demand on price mark downs.
<br>
**Features Created**
<br>
`is_mark_down{1-5}_na`
<br>
`total_mark_down`
<br>
`total_mark_down_intensity`
<br>
`mark_down{1-5}_intensity`

```{r mark_down_fts, echo=TRUE, message=FALSE, warning=FALSE}

#====== Create mark down features ======

mark_down_fts_tbl <- base_tbl %>%
  select(store, dept, week_dates, starts_with("mark_down")) %>%
  left_join(lag_demand_fts_tbl %>%
              select(store, dept, week_dates, lag_weekly_sales_1),
            by = c("store", "dept", "week_dates")
            ) %>%
  mutate(
    
##NA in this data at times means no mark down and at other times means missing data
##to differentiate the two and extract signals I create NA features

    #---- mark down NA features ----
    across(starts_with("mark_down"),
           ~ is.na(.x),
           .names = "is_{.col}_na"),
    
    #---- total mark downs in a store department feature ----
    total_mark_down = rowSums(
      across(starts_with("mark_down")),
      na.rm = TRUE),
    
    #---- total mark down relative to store department demand feature ----
    total_intensity_mark_down = total_mark_down / pmax(lag_weekly_sales_1, 1),

    #---- individual mark downs relative to store department demand features ----
    across(starts_with("mark_down"),
      ~ .x / pmax(lag_weekly_sales_1, 1),
      .names = "intensity_{.col}")
    
    ) %>%
    select(store, dept, week_dates, contains(c("is", "total", "intensity")))

```
<br>

#### Feature Assembly
**Purpose:** To join pre split engineered features into a singular table.

```{r ft_assembly, echo=TRUE, message=FALSE, warning=FALSE}

#====== Categorize feature layers for joining ======

global_ft_layers <- list(calendar_fts_tbl, 
                         holiday_event_fts_tbl, 
                         lead_holiday_fts_tbl, 
                         lag_holiday_fts_tbl, 
                         economic_fts_tbl)


cross_sectional_ft_layers <- list(lag_demand_fts_tbl, 
                                  level_demand_fts_tbl, 
                                  variability_demand_fts_tbl, 
                                  mark_down_fts_tbl, 
                                  momentum_demand_fts_tbl)





#====== Join all feature layers ======
  
global_ft_tbl <- global_ft_layers %>%
  reduce(left_join,
         by = "week_dates")


cross_sectional_ft_tbl <- cross_sectional_ft_layers %>%
  reduce(left_join,
         by = c("store", "dept", "week_dates"))



assembled_fts_tbl <- base_tbl %>%
  left_join(global_ft_tbl,
            by = "week_dates") %>%
  left_join(cross_sectional_ft_tbl,
            by = c("store", "dept", "week_dates"))





#====== Display tbl ======

glimpse(assembled_fts_tbl)

```
<br>

#### Train/Test
**Purpose:** To establish a true out of sample evaluation period that simulates future forecasting.

```{r train_test, echo=TRUE, message=FALSE, warning=FALSE}

#====== Split into train data and test data ======

##the data is from 2010-02-05 to 2012-11-01, there is a approximately an 80/20 split after date grouping is taken into consideration

train_tbl <- assembled_fts_tbl %>% 
  filter(week_dates < "2012-04-06") 

test_tbl <- assembled_fts_tbl %>%
  filter(week_dates >= "2012-04-06")

```
<br>

#### Relative Demand Features
**Purpose:** To normalize sales relative to store and department historical averages to stabilize scale differences across entities.
<br>
**Features Created**
<br>
`{store_dept, store,dept}_avg`
<br>
`vs_{store_dept, store,dept}`

```{r relative_demand_fts, echo=TRUE, message=FALSE, warning=FALSE}

#====== Compute store department sales baselines ======

avg_store_dept_tbl <- train_tbl %>%
  select(store, dept, weekly_sales) %>%
  group_by(store, dept) %>%
  summarise(
    store_dept_avg = mean(weekly_sales, na.rm = TRUE),
    .groups = "drop"
  )





#====== Compute store sales baselines ======

avg_store_tbl <- train_tbl %>%
  select(store, weekly_sales) %>%
  group_by(store) %>%
  summarise(
    store_avg = mean(weekly_sales, na.rm = TRUE),
    .groups = "drop"
    )





#====== Compute dept sales baselines ======

avg_dept_tbl <- train_tbl %>%
  select(dept, weekly_sales) %>%
  group_by(dept) %>%
  summarise(
    dept_avg = mean(weekly_sales, na.rm = TRUE),
    .groups = "drop"
    )





#====== Create relative demand features in train data and test data ======

relative_train_demand_fts_tbl <- train_tbl %>%
  select(store, dept, week_dates, lag_weekly_sales_1) %>%
  left_join(avg_store_dept_tbl,
            by = c("store","dept")) %>%
  left_join(avg_store_tbl,
            by = "store") %>%
  left_join(avg_dept_tbl,
            by = "dept") %>%
  mutate(
    
    vs_store_dept  = lag_weekly_sales_1 / pmax(coalesce(store_dept_avg, 1), 1),
    vs_store       = lag_weekly_sales_1 / pmax(coalesce(store_avg, 1), 1),
    vs_dept        = lag_weekly_sales_1 / pmax(coalesce(dept_avg, 1), 1)
    
  ) %>%
  select(store, dept, week_dates, contains(c("vs", "avg")))



relative_test_demand_fts_tbl <- test_tbl %>%
  select(store, dept, week_dates, lag_weekly_sales_1) %>%
  left_join(avg_store_dept_tbl,
            by = c("store","dept")) %>%
  left_join(avg_store_tbl,
            by = "store") %>%
  left_join(avg_dept_tbl,
            by = "dept") %>%
  mutate(
    
    vs_store_dept  = lag_weekly_sales_1 / pmax(coalesce(store_dept_avg, 1), 1),
    vs_store       = lag_weekly_sales_1 / pmax(coalesce(store_avg, 1), 1),
    vs_dept        = lag_weekly_sales_1 / pmax(coalesce(dept_avg, 1), 1)

  ) %>%
  select(store, dept, week_dates, contains(c("vs", "avg")))





#====== Join relative demand features to train data and test data ======

train_model_input_tbl <- train_tbl %>%
  left_join(relative_train_demand_fts_tbl,
            by = c("store", "dept", "week_dates"))
  

test_model_input_tbl <- test_tbl %>%
  left_join(relative_test_demand_fts_tbl,
            by = c("store", "dept", "week_dates"))

```
<br>

#### Model Framework
**Purpose:** Define preprocessing, model structure, and workflow abstraction for reproducible training and tuning.

```{r model_framework, echo=TRUE, message=FALSE, warning=FALSE}

#====== Define preprocessing for train data and test data ======

sales_recipe <- recipe(weekly_sales ~ ., data = train_model_input_tbl) %>%
  update_role(dept, week_dates, store, store_dept, new_role = "id") %>%
  step_mutate(
    
    #---- replace NA in mark_down columns with 0 ----
    across(starts_with("mark_down"), ~ replace_na(.x, 0)),
    
    #---- convert logical features into integer features ----
    across(where(is.logical), as.integer)
    
    ) %>%

  #---- impute remaining numeric NAs (lags, rolls) ----
  step_impute_median(all_numeric_predictors()) %>%

  #---- dummify all nominal predictors ----
  step_dummy(all_nominal_predictors())





#====== Define a boosted tree model (XGBoost) ======

initial_model <- boost_tree(
  
  trees          = tune(),
  tree_depth     = tune(),
  learn_rate     = tune(),
  min_n          = tune(),
  loss_reduction = tune(),
  sample_size    = tune()
  
  ) %>%
  set_engine("xgboost", 
             seed = 2067205,
             objective = "reg:squarederror") %>% ##squared error as loss function
  set_mode("regression") 





#====== Define a workflow for initial model ======

initial_workflow <- workflow() %>%
  add_model(initial_model) %>%
  add_recipe(sales_recipe)

```
<br>

#### Processed Data Validation
**Purpose:** To ensure the processed data is suitable for modeling.

```{r processed_data_validation, message=FALSE, warning=FALSE, include=FALSE}

#====== Apply preprocessing to train data ======

prepped_sales_recipe <- prep(sales_recipe, training = train_model_input_tbl, retain = TRUE) 

baked_train_model_input_tbl <- bake(prepped_sales_recipe, new_data = NULL)





#====== Inspect data ======

glimpse(baked_train_model_input_tbl)





#====== Check for missing values ====

colSums(is.na(baked_train_model_input_tbl))

##Result is that there are no missing values 





#====== Check for infinite values ======

colSums(sapply(baked_train_model_input_tbl, is.infinite))

##Result is that there are no infinite values





#====== Check number of stores and departments =====

n_distinct(baked_train_model_input_tbl$store)


n_distinct(baked_train_model_input_tbl$dept)

##There are 45 stores and 81 departments





#====== Check statistics for key numeric variables =====

baked_train_model_input_tbl %>%
  select(store, dept, weekly_sales) %>%
  summary()

##All store and dept ids are positive





#====== Check for duplicates ======

anyDuplicated(baked_train_model_input_tbl)

##Result is that there are no duplicated rows

```
<br>

#### Cross Validation
**Purpose:** Evaluate model stability across multiple historical forecasting scenarios using expanding window cross validation.

```{r cross_validation, echo=TRUE, message=FALSE, warning=FALSE}

##the rolling_origin() function by default treats each row as a separate period, but the Walmart sales data set has multiple rows per period
##to solve this I use the rolling_origin function on distinct Walmart sales weeks, then I map the splits back to the full train data

#====== Create a tbl with all distinct weeks in chronological order ======

weeks_tbl <- train_model_input_tbl %>%
  distinct(week_dates) %>%
  arrange(week_dates)

#nrow(week_tbl) 
##114 weeks total





#====== Define expanding window cross validation ======

week_folds <- rolling_origin(
  data       = weeks_tbl,
  initial    = 78,    ##trains on 1.5 years since only 1 year is high variance
  assess     = 4, 
  skip       = 0,
  cumulative = TRUE
)

##there are 33 folds





#====== Define a function to map splits to a data set ======

split_function <- function(split, data) {
  

  #---- get the analysis and assessment weeks ---- 
  analysis_weeks  <- analysis(split)$week_dates 
  assessment_weeks <- assessment(split)$week_dates 
  
  #---- map weeks back to the train data set ----
  analysis_rows <- which(data$week_dates %in% analysis_weeks)
  assessment_rows <- which(data$week_dates %in% assessment_weeks)
  
  #---- turn into a split object ----
  make_splits(
    list(
      analysis = analysis_rows,
      assessment = assessment_rows
      ),
    data
    )
  
}
  
  
  
  
  
#====== Map rolling_origin() splits to train data ======

sales_splits <- map(week_folds$splits, split_function, train_model_input_tbl)

sales_folds <- manual_rset(sales_splits, ids = week_folds$id) ##turns splits list into resampling set

```
<br>

#### Weighted Average Percentage Error
**Purpose:** Define and implement WAPE as the primary evaluation metric to measure relative forecast error scaled by total sales volume, ensuring interpretability across heterogeneous retail demand levels.

```{r performance_metric, echo=TRUE, message=FALSE, warning=FALSE}

##The code below defines a custom weighted absolute percentage error (WAPE) metric in the style recommended by the yardstick package developers. See https://www.tidymodels.org/learn/develop/metrics/ for more information on yardstick developer functions.

#====== Define a WAPE implementation function ======

wape_impl <- function(truth, estimate, case_weights = NULL){
  
  denominator <- sum(abs(truth), na.rm = TRUE) ##make function defensively robust to NA
  
  if (denominator == 0) {
    return(NA_real_)
  }
  
  sum(abs(truth - estimate)) / denominator
}





#====== Define a vector implementation of the WAPE function ======

wape_vec <- function(truth, estimate, na_rm = TRUE, case_weights = NULL, ...) {##new metric functions must have the first 4 arguments to be compatible with the yardstick package
 
  check_numeric_metric(truth, estimate, case_weights) ##verifies that all arguments are numeric

  if (na_rm) {
    
    result <- yardstick_remove_missing(truth, estimate, case_weights) ##removes rows where any argument is missing
    
    ##creates new vectors of clean values
    truth <- result$truth
    estimate <- result$estimate
    case_weights <- result$case_weights
    
    
  } else if (yardstick_any_missing(truth, estimate, case_weights)) { ##returns TRUE if any argument missing
 
    return(NA_real_)
    
  }
  
  wape_impl(truth, estimate, case_weights = case_weights)
}





#====== Define a data frame implementation of the WAPE function ======

wape <- function(data, ...) {
  UseMethod("wape") ##decides which specific function to call based on the class of the object provided
} 


wape.data.frame <- function(data, truth, estimate, na_rm = TRUE, case_weights = NULL, ...) {

  numeric_metric_summarizer(
    name = "wape",
    fn = wape_vec,
    data = data,
    truth = !!enquo(truth), 
    estimate = !!enquo(estimate),
    na_rm = na_rm,
    case_weights = !!enquo(case_weights)
  )
  
}





#====== Add WAPE metric ======

wape <- new_numeric_metric(wape, "minimize")

##wape metric has the same class as yardstick base metrics and can now be used for tuning

sales_metrics <- metric_set(wape)

```
<br>

#### Broad Tuning
**Purpose:** To search a broad parameter space for a plausible region that produces stable models.

```{r broad_tune, echo=TRUE, message=FALSE, warning=FALSE}

#====== Load or initiate broad tune ======

version <- "v1"
broad_tuning_path <- paste0("r/broad_tune_", version, ".rds")


if (file.exists(broad_tuning_path)) {
  
  message("Loading existing broad tune...")
  broad_tune <- readRDS(broad_tuning_path)
  
} else {

  
  
  
  
##Parameter ranges below were selected using practical heuristics from gradient boosting literature and prior experience

#====== Define a broad range of possible tuning parameters ======

broad_grid <- grid_space_filling(
  
  trees(
    range = c(500, 1800)
    ),
  tree_depth(
    range = c(1, 8)
    ),
  learn_rate(
    range = c(-4, -1) ##ranges are specified on a log10 scale
    ), 
  min_n(
    range = c(5, 45)
  ),
  loss_reduction(
    range = c(-8, -1)
  ),
  sample_size = sample_prop(
    range = c(0.2, 0.7)
  ),
  size = 30
  
  )





#====== Broadly tune parameters ======

broad_tune <- tune_race_anova( ##tune_race_anova() eliminates poorly performing parameter combinations early
  initial_workflow,
  resamples = sales_folds,
  grid      = broad_grid,
  metrics   = sales_metrics,
  control   = control_race(
    save_pred = TRUE,
    verbose   = TRUE
    )
)





#====== Save broad tune ======

saveRDS(broad_tune, broad_tuning_path)

}





#====== Display best broad models ======

best_broad_models  <- show_best(broad_tune, metric = "wape", n = 5)

##while the show_best() function has argument n = 5 specifying to show top 5 models, tune_race_anova() function eliminates models early on, explaining why there are less than 5 models shown in output

best_broad_models


best_broad_model <- select_best(broad_tune, metric = "wape")

```
<br>

#### Fine Tuning
**Purpose:** To search a reduced parameter space for a plausible region that produces stable models.

```{r fine_tune, echo=TRUE, message=FALSE, warning=FALSE}

#====== Load or initiate fine tune ======

version <- "v1"
fine_tuning_path <- paste0("r/fine_tune_", version, ".rds")


if (file.exists(fine_tuning_path)) {
  
  message("Loading existing fine tune...")
  fine_tune <- readRDS(fine_tuning_path)
  
} else {

  
  
  

##Parameter ranges below were selected using practical heuristics from gradient boosting literature and prior experience

#====== Define ranges close to the best broad tuning parameters for fine tuning ======

fine_grid <- grid_space_filling(
  
  trees(range = c(
    as.integer(best_broad_model$trees * 0.75),
    as.integer(best_broad_model$trees * 1.25)
  )),

  tree_depth(range = c(
    max(2, best_broad_model$tree_depth - 1), ##keep minimum at 2 so tuning stays within a useful region of the parameter space
    best_broad_model$tree_depth + 1
  )),

  learn_rate(range = log10(c(
    best_broad_model$learn_rate * 0.7,
    best_broad_model$learn_rate * 1.3
  ))),

  min_n(range = c(
    max(7, best_broad_model$min_n - 7),
    best_broad_model$min_n + 7
  )),

  loss_reduction(range = log10(c(
    best_broad_model$loss_reduction * 0.3,
    best_broad_model$loss_reduction * 3
  ))),

  sample_size = sample_prop(range = c(
    max(0.5, best_broad_model$sample_size - 0.1),
    min(0.95, best_broad_model$sample_size + 0.1)
  )),

  size = 20
)





#====== Finely tune parameters ======

fine_tune <- tune_race_anova( ##tune_race_anova() eliminates poorly performing parameter combinations early
  initial_workflow,
  resamples = sales_folds,
  grid      = fine_grid,
  metrics   = sales_metrics,
  control   = control_race(
    save_pred = TRUE,
    verbose   = TRUE
    )
)





#====== Save fine tune ======

saveRDS(fine_tune, fine_tuning_path)

}





#====== Display best fine tune models ======

best_fine_models  <- show_best(fine_tune, metric = "wape", n = 5)

##while the show_best() function has argument n = 5 specifying to show top 5 models, tune_race_anova() function eliminates models early on, explaining why there are less than 5 models shown in output

best_fine_models 


best_fine_model <- select_best(fine_tune, metric = "wape")

```
<br>

#### Final Model Training
**Purpose:** To train the final model using the full training data.

```{r train_final_model, echo=TRUE, message=FALSE, warning=FALSE}

#====== Define final model using full train data ======

final_workflow <- finalize_workflow(initial_workflow, best_fine_model)





#====== Train model using full train data ======

final_model_fit <- fit(final_workflow, data = train_model_input_tbl)

```
<br>

#### Final Model Evaluation
**Purpose:** To evaluate out of sample performance on the strictly held out test data using WAPE.

```{r final_model, echo=TRUE, message=FALSE, warning=FALSE}

#====== Predict test data using final model ======

final_test_predictions <- predict(final_model_fit, new_data = test_model_input_tbl) %>%
  bind_cols(test_model_input_tbl %>% select(dept, weekly_sales))





#====== Evaluate model performance at department level ======

dept_model_evaluation <- final_test_predictions %>%
  group_by(dept) %>%
  wape(truth = weekly_sales, estimate = .pred)





#====== Evaluate final model performance overall ======

final_model_evaluation <- final_test_predictions %>%
  wape(truth = weekly_sales, estimate = .pred)

final_model_evaluation
```
<br>

#### Model Interpetation
**Purpose:** To clearly visualize overall forecast model performance.

```{r model_interpretation, echo=TRUE, message=FALSE, warning=FALSE}

#====== Plot final model feature importance ======

feature_importance_plot <- extract_fit_engine(final_model_fit) %>% 
  vip(num_features = 5) + ##top 5 features
  
  geom_col(fill = "#1F78B4") +
  
  #---- add titles and axis labels ----
  labs(
    title = "Top 5 Most Important Features",
    subtitle = "XGBoost Feature Importance",
    x = "Feature",
    y = "Importance"
    ) +
  
  #---- aesthetic changes to plot ----
  theme_minimal(base_size = 11) +
  theme(
    
    panel.grid = element_blank(), 
    
    plot.title = element_text(hjust = 0.5, face = "bold", size = 17), 
    
    plot.subtitle = element_text(hjust = 0.5, size = 13),
    
    axis.line = element_line(color = "black", linewidth = 0.4), 
    
    legend.position = "none" 
    
    )

feature_importance_plot





#====== Aggregate true and estimated demand to total weekly level ======

predictions_plot_tbl <- final_test_predictions %>%
  ##can safely bind week_dates since predict() function preserves order of original test data
  bind_cols(test_tbl %>% select(week_dates)) %>% 
  group_by(week_dates) %>%
  summarise(
    
    truth = sum(weekly_sales),
    estimation = sum(.pred),
    .groups = "drop"
    
  )





#====== Filter out extreme department WAPE values ======

dept_model_evaluation_filtered <- dept_model_evaluation %>%
  filter(.estimate < 1)





#====== Plot distribution of department WAPE ======

dept_wape_distribution_plot <- ggplot(dept_model_evaluation_filtered, aes(x = .estimate)) +

  #---- specify histogram ----
  geom_histogram(
    bins = 12,
    fill = "#1F78B4",
    color = "black",
    alpha = 0.85
  ) +

  #---- add titles and labels ----
  labs(
    title = "Distribution of Department Forecast Error",
    subtitle = "Department level forecast accuracy across Walmart sales",
    x = "Department WAPE",
    y = "Count of Departments"
  ) +

  #---- format x axis as percentages ----
  scale_x_continuous(labels = percent_format(accuracy = 1)) +

  #---- aesthetic changes to plot ----
  theme_minimal(base_size = 11) +

  theme(
    
    panel.grid = element_blank(),
    
    plot.title = element_text( 
      hjust = 0.5,
      face = "bold",
      size = 17
    ),
    
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 12
    ),
    
    axis.line = element_line( 
      color = "black",
      linewidth = 0.4
    )
    
    )

dept_wape_distribution_plot





#====== Save distribution of department WAPE plot ======

ggsave("dept_wape_distribution_plot.png", dept_wape_distribution_plot, width = 10, height = 6, dpi = 200)





#====== Pull most difficult departments to forecast ======

difficult_depts_tbl <- dept_model_evaluation %>%
  arrange(desc(.estimate)) %>%
  slice_head(n = 5) %>%
  mutate(dept = reorder(as.factor(dept), .estimate))





#====== Plot most difficult departments to forecast ======

difficult_depts_plot <- ggplot(difficult_depts_tbl, aes(x = dept, y = .estimate)) +
  
  geom_col(
    fill = "#1F78B4",
    width = 0.7
  ) +
  
  coord_flip() + #swap x axis and y axis
  
  #---- add titles and labels ----
  labs(
    title = "Top 5 Departments with Highest Forecast Error",
    subtitle = "Departments with the largest out of sample WAPE",
    x = "Department Number",
    y = "Department WAPE"
  ) +
  
  #---- aesthetic changes ----
  theme_minimal(base_size = 11) +
  
  theme(
    
    panel.grid = element_blank(),
    
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 17),
      
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 13
    ),
    
    axis.line = element_line(
      color = "black",
      linewidth = 0.4
    )
    
  )

difficult_depts_plot





#====== Plot true vs estimated demand ======

predictions_plot <- ggplot(predictions_plot_tbl, aes(x = week_dates)) +
  
  #---- add truth and prediction lines on plot ----
  geom_line(aes(y = truth, color = "Actual Sales"), linewidth = 0.9) + 
  geom_line(aes(y = estimation, color = "Model Prediction"), linewidth = 0.7) +
  
  #---- add titles and axis labels ----
  labs( 
    title = "Walmart Retail Sales Forecast",
    subtitle = "Test set performance - WAPE = 8.4%",
    y = "Total Weekly Sales (Millions USD)",
    x = "Month",
    color = NULL 
  ) +
  
  #---- specify line colors ----
  scale_color_manual(
    values = c(
    "Actual Sales" = "black",
    "Model Prediction" = "#1F78B4"
    )
    ) +

  #---- transformations to y axis values ----
  scale_y_continuous(
    labels = label_dollar(scale = 1e-6, accuracy = 1, suffix = "M"), 
    expand = expansion(mult = c(0.1, 0.2))
    ) +
  
  #---- transformations to x axis values ----
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +

  #---- aesthetic changes to plot ----
  theme_minimal(base_size = 11) +
  theme(
    
    panel.grid = element_blank(),
    
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 17
      ),
    
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 13
      ),
    
    axis.line = element_line(
      color = "black",
      linewidth = 0.4
      ),
    
    axis.title.y = element_text(margin = margin(r = 12)), 
    
    legend.position = "top" 
    
    )


predictions_plot





#====== Save predictions plot ======

ggsave("predictions_plot.png", predictions_plot, width = 10, height = 6, dpi = 200)

```
